Hierarchical Principal Manifold Estimation

### 2023-08-15 paper/main.tex
add beginning of varying-coefficient estimation algorithm

We begin this section with a brief review of additive models. Consider a regression scenario where output variable $\mathbf{y} = \left(y_1, \dots, y_n\right)^{T}$ are thought to be influenced by covariate vectors $\mathbf{x}_i = \left(x_{i1}, \dots, x_{ip}\right)$. In this situation, the standard multiple linear regression model assumes that the dependence of $y_i$ on $\mathbf{x}_i$ takes the linear form
\[%
  y_i = \alpha + x_{i1}\beta_1 + \dots + x_{ip}\beta_p + \epsilon_i
,\]%
where $\epsilon$ has mean 0 and variance $\sigma^2$. Additive models are a direct extension of this model, where the additive structure of the model is retained, but the assumption of linearity is discarded. A simple additive model takes the form
\[%
  y_i = \alpha + \sum_{j=1}^{p}f_j(X_{ij}) + \epsilon_i
,\]%
where again $\epsilon_i$ has mean 0 and variance $\sigma^2$, but the $f_j$s represent arbitrary functions which are commonly smooth. While the functions in the model above only take as parameters the value of one covariate, with each covariate being represented in the model once, this does not need to be the case. The structure of additive models seeks to balance benefits of interpretability found in additive models with the flexibility of nonlinear, and frequently, nonparametric functional forms. Further information can be found in \cite{hastieGeneralizedAdditiveModels1990}.

@@

Individual participants are typically scheduled for imaging at six month intervals, though variability exists in terms of the recorded intervals. Also of note, the numbers of observations contributed by each participant varies, so groups may be unbalanced. We denote the total number of participants as $N$, and use $n_i$ to represent the number of images present for each individual. Meanwhile, $G$ denotes the number of observed groups under consideration.

Within a segmented structural MRI, the voxels estimated to mark the surface of the hippocampus can be assumed to lie along a two-dimensional manifold that has been wrapped into the three-dimensional image space. More generally, we let $d$ denote the intrinsic dimension of the low-dimensional manifold, and $D$ represent the dimension of the high-dimensional input space. For the $j$th image from the $i$th individual belonging to group $g$ with $i = 1, \dots, N$, $j = 1, \dots, n_i$, and $g = 1, \dots, G$, we denote the $k$th voxel in the high-dimensional space as $\mathbf{x}_{ijk} \in \mathbb{R}^{3}$, and its associated low-dimensional parameterization as $\mathbf{y}_{ijk} \in \mathbb{R}^2$ and weight $w_{ijk}$. We let $t_{ij}$ denote the time at which the $j$th image for individual $i$ is recorded. The embedding function which maps from the $d$-dimensional manifold to $D$-dimensional ambient space is represented as $\psi(\mathbf{y}_{ijk})$, which can be modeled as \[%
  \mathbf{x}_{ijk} = \psi(\mathbf{y}_{ijk}) = f_p(\mathbf{y}_{ijk}, t_{ij}) + f_{g(i)}(\mathbf{y}_{ijk}, t_{ij}) + f_i(\mathbf{y}_{ijk}, t_{ij}) + f_{ij}(\mathbf{y}_{ijk}) + \epsilon
.\]%

Each function in this additive model represents a different level of nesting in the data, with $f_p(\mathbf{y}, t)$ estimating the embedding at the population level, and $f_g(\mathbf{y}, t)$, $f_i(\mathbf{y}, t)$, and $f_{ij}(\mathbf{y})$ describing deviations from the population estimates specific to the group, individual, and image in question, respectively. As estimating longitudinal changes in the underlying manifold may be of interest, $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$ and $f_i(\mathbf{y}, t)$ are each denoted as varying-coefficient models (\cite{hastieVaryingCoefficientModels1993}). Thus, the model is capable of estimating separate longitudinal changes for population, group, and individual estimates. In cases where longitudinal changes are not of interest, each varying-coefficient model may be constrained to be held constant with respect to time. As the $ij$th image is recorded at a single time point, $f_{ij}(\mathbf{y}, t)$ may not be dependent on time. We specify $f_{ij}(\mathbf{y})$ as a smoothing spline, while the functions $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$, and $f_i(\mathbf{y}, t)$ are each varying-coefficient smoothing spline models, estimated by using a smoothing spline to smooth over smoothing spline coefficients calculated at each time point available in the data.

The smoothing spline functions $f_{ij}$ minimize 
\[%
  \mathcal{K}_{\lambda}(f_{ij}) = \sum_{l=1}^{D}\left\{\sum_{k=1}^{m_{ij}}w_{ijk}|\mathbf{x}_{ijk, l} - f_{ij, l}(\mathbf{y}_{ijk})|^2 + \lambda\|\nabla^{\bigotimes 2}f_l\|^2\right\}
,\]%
where $\nabla^{\bigotimes 2}f$ denotes the Hessian matrix of $f$, and $\lambda$ is given. Thus, $f_{ij}$ takes the form
\[%
  f_{ij, l}(\mathbf{y}) = \sum_{k=1}^{m_{ij}} s_{k, l} \times \eta_{4-d}\left(\mathbf{y} - \mathbf{y}_{ijk}\right) + \sum_{p=1}^{d+1} \alpha_{p, l} \times \phi_p(\mathbf{y}), \quad l = 1, 2, \dots, D
,\]%
with $\phi_p(\mathbf{y})$ representing the basis of the linear space of polynomials on $\mathbb{R}^{d}$ with degree $\leq 1$ and $\eta_{\nu}(\mathbf{y}) = \|\mathbf{y}\|_{\mathbb{R}^{d}}^{\nu}\log\left(\|\mathbf{y}\|_{\mathbb{R}^{d}}\right)$ when $\|\mathbf{y}\|_{\mathbb{R}^{d}} \neq 0$ and $\nu$ is even, $\eta_{\nu}(\mathbf{y}) = 0$ when $\|\mathbf{y}\|_{\mathbb{R}^{d}} = 0$ and $\nu$ is even, and $\eta_{\nu}(\mathbf{y}) = \|\mathbf{y}\|_{\mathbb{R}^{d}}^{\nu}$ when $\nu$ is odd.

@@

\[%
  \left(\begin{array}{ccc}
    2\mathbf{EWE} + 2\lambda\mathbf{E} & 2\mathbf{EWR} & \mathbf{R} \\
    2\mathbf{R}^{T}\mathbf{WE} & 2\mathbf{R}^{T}\mathbf{WR} & \mathbf{0} \\
    \mathbf{R}^{T} & \mathbf{0} & \mathbf{0}
  \end{array}\right)\left(
  \begin{array}{c}
    s_l \\
    \alpha_l \\
    m_l
  \end{array}
  \right) = \left(
  \begin{array}{c}
    2\mathbf{EW}\mathbf{x}_l \\
    2\mathbf{R}^{T}\mathbf{W}\mathbf{x}_l \\
    \mathbf{0}
  \end{array}
  \right)
.\]%

>>>>

We begin this section with a brief review of additive models. Consider a regression scenario where output variable $\mathbf{y} = \left(y_1, \dots, y_n\right)^{T}$ are thought to be influenced by covariate vectors $\mathbf{x}_i = \left(x_{i1}, \dots, x_{ip}\right)$. In this situation, the standard multiple linear regression model assumes that the dependence of $y_i$ on $\mathbf{x}_i$ takes the linear form
\[%
  y_i = \alpha + x_{i1}\beta_1 + \dots + x_{ip}\beta_p + \epsilon_i
,\]%
where $\epsilon$ has mean 0 and variance $\sigma^2$. Additive models are a direct extension of this model, where the additive structure of the model is retained, but the assumption of linearity is discarded. A simple additive model takes the form
\begin{equation}
  y_i = \alpha + \sum_{j=1}^{p}f_j(X_{ij}) + \epsilon_i, \label{eq:1}
\end{equation}
where again $\epsilon_i$ has mean 0 and variance $\sigma^2$, but the $f_j$s represent arbitrary functions which are commonly smooth. While the functions in model \ref{eq:1} only take as parameters the value of one covariate, with each covariate being represented in the model once, this does not need to be the case. The structure of additive models seeks to balance benefits of interpretability found in additive models with the flexibility of nonlinear, and frequently, nonparametric functional forms. Further information can be found in \cite{hastieGeneralizedAdditiveModels1990}.

@@

Individual participants are typically scheduled for imaging at six month intervals, though variability exists in terms of the recorded intervals. Also of note, the numbers of observations contributed by each participant varies, so groups may be unbalanced. We use $G$ to denote the number of observed groups under consideration, then denote the number of individuals as $N$, the number of images belonging to the $i$th individual as $n_i$, and the number of voxels from the $j$th image of individual $i$ as $m_{ij}$.

Within a segmented structural MRI, the voxels estimated to mark the surface of the hippocampus can be assumed to lie along a two-dimensional manifold that has been wrapped into the three-dimensional image space. More generally, we let $d$ denote the intrinsic dimension of the low-dimensional manifold, and $D$ represent the dimension of the high-dimensional input space. For the $j$th image from the $i$th individual belonging to group $g$ with $i = 1, \dots, N_g$, $j = 1, \dots, n_i$, and $g = 1, \dots, G$, we denote the $k$th voxel in the $D$-dimensional space as $\mathbf{x}_{ijk} \in \mathbb{R}^{D}$, its associated $d$-dimensional parameterization as $\mathbf{y}_{ijk} \in \mathbb{R}^{d}$, and its weight as $w_{ijk}$. We let $t_{ij}$ denote the time at which the $j$th image for individual $i$ is recorded, with all times measured as the difference between the date the image was taken and the date of the participant's baseline visit. We use $\mathbf{x}_g$, $\mathbf{x}_i$, $\mathbf{x}_{ij}$, and $\mathbf{x}_t$ to denote the input voxels from all images in the $g$th group, $i$th individual, $ij$th image, and $t$th time point, respectively, and use similar notation for all other relevant terms.

The embedding function which maps from the $d$-dimensional manifold to $D$-dimensional ambient space is represented as $\psi(\mathbf{y}_{ijk})$, which can be modeled as 
\begin{equation}
  \mathbf{x}_{ijk} = \psi(\mathbf{y}_{ijk}) = f_p(\mathbf{y}_{ijk}, t_{ij}) + f_{g(i)}(\mathbf{y}_{ijk}, t_{ij}) + f_i(\mathbf{y}_{ijk}, t_{ij}) + f_{ij}(\mathbf{y}_{ijk}) + \epsilon. \label{eq:2}
\end{equation}
Each function in this additive model represents a different level of nesting in the data, with $f_p(\mathbf{y}, t)$ estimating the embedding at the population level, and $f_g(\mathbf{y}, t)$, $f_i(\mathbf{y}, t)$, and $f_{ij}(\mathbf{y})$ describing deviations from the population estimates specific to the group, individual, and image in question, respectively. As estimating longitudinal changes in the underlying manifold may be of interest, $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$ and $f_i(\mathbf{y}, t)$ are each denoted as varying-coefficient models, where the coefficients of each function are dependent on time (\cite{hastieVaryingCoefficientModels1993}). Thus, the model is capable of estimating separate longitudinal changes for population, group, and individual estimates. In cases where longitudinal changes are not of interest, each varying-coefficient model may be constrained to be held constant with respect to time. As the $ij$th image is recorded at a single time point, $f_{ij}(\mathbf{y}, t)$ may not be dependent on time. We specify $f_{ij}(\mathbf{y})$ as a smoothing spline, while the functions $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$, and $f_i(\mathbf{y}, t)$ are each varying-coefficient smoothing spline models, estimated by using a smoothing spline to smooth over smoothing spline coefficients calculated at each time point available in the data.

The smoothing spline functions $f_{ij}$ minimize 
\begin{equation}
  \mathcal{K}_{\lambda}(f_{ij}) = \sum_{l=1}^{D}\left\{\sum_{k=1}^{m_{ij}}w_{ijk}|\mathbf{x}_{ijk, l} - f_{ij, l}(\mathbf{y}_{ijk})|^2 + \lambda\|\nabla^{\otimes 2}f_l\|^2\right\}, \label{eq:3}
\end{equation}
where $\nabla^{\otimes 2}f$ denotes the Hessian matrix of $f$, and $\lambda$ is given. Thus, $f_{ij}$ takes the form
\begin{equation}
  f_{ij, l}(\mathbf{y}) = \sum_{k=1}^{m_{ij}} s_{k, l} \times \eta_{4-d}\left(\mathbf{y} - \mathbf{y}_{ijk}\right) + \sum_{p=1}^{d+1} \alpha_{p, l} \times \phi_p(\mathbf{y}), \quad l = 1, 2, \dots, D, \label{eq:4}
\end{equation}
with $\phi_p(\mathbf{y})$ representing the basis of the linear space of polynomials on $\mathbb{R}^{d}$ with degree $\leq 1$ and $\eta_{\nu}(\mathbf{y}) = \|\mathbf{y}\|_{\mathbb{R}^{d}}^{\nu}\log\left(\|\mathbf{y}\|_{\mathbb{R}^{d}}\right)$ when $\|\mathbf{y}\|_{\mathbb{R}^{d}} \neq 0$ and $\nu$ is even, $\eta_{\nu}(\mathbf{y}) = 0$ when $\|\mathbf{y}\|_{\mathbb{R}^{d}} = 0$ and $\nu$ is even, and $\eta_{\nu}(\mathbf{y}) = \|\mathbf{y}\|_{\mathbb{R}^{d}}^{\nu}$ when $\nu$ is odd.

@@

\begin{equation}
   \left(\begin{array}{ccc}
    2\mathbf{EWE} + 2\lambda\mathbf{E} & 2\mathbf{EWR} & \mathbf{R} \\
    2\mathbf{R}^{T}\mathbf{WE} & 2\mathbf{R}^{T}\mathbf{WR} & \mathbf{0} \\
    \mathbf{R}^{T} & \mathbf{0} & \mathbf{0}
  \end{array}\right)\left(
  \begin{array}{c}
    s_l \\
    \alpha_l \\
    m_l
  \end{array}
  \right) = \left(
  \begin{array}{c}
    2\mathbf{EW}\mathbf{x}_l \\
    2\mathbf{R}^{T}\mathbf{W}\mathbf{x}_l \\
    \mathbf{0}
  \end{array}
  \right). \label{eq:5}
\end{equation}
The optimal value of $\lambda$ can be found using cross-validation, with the functions being evaluated by the mean squared error, written as
\begin{equation}
  MSE(\lambda) = \frac{1}{m_{ij}}\sum_{k=1}^{m_{ij}}\left(\mathbf{x}_{ijk} - f_{ij}(\mathbf{y}_{ijk})\right)^2. \label{eq:6}
\end{equation}

The solutions to this minimization problem suffice for estimating $f_{ij}(\mathbf{y})$, in which the coefficients are not time-varying. To reach estimates for the varying-coefficient models, equation \ref{eq:5} is solved using the data available at each time point within the relevant group. To estimate $f_p(\mathbf{y}, t)$, we use $\left\{\mathbf{x}_t\right\}_{t=1}^T$ and $\left\{\mathbf{y}_t\right\}_{t=1}^T$, yielding fitted function $\left\{\hat{f}_t(\mathbf{y})\right\}_{t=1}^T$. Using the maximum and minimum values of $\mathbf{y}_t$ in each dimension, a new $d$-dimensional grid of $m' = \text{max}\left(\left\{m_{ij}\right\}_{i=1, j = 1}^{N, N_i}\right)$ knots, $\left\{\mathbf{y}'\right\}$, is generated with the knots being evenly spread throughout the parameter space. These parameters serve as common knots that allow the spline coefficients to be compared across time points. They are then used to compute a new set of corresponding outcome values $\left\{\mathbf{x}'_t\right\}_{t=1}^{T}$. The spline functions are reestimated as $\left\{\hat{f}'_t\right\}_{t=1}^{T}$ using $\left\{\mathbf{x}'_t\right\}_{t=1}^{T}$ and $\mathbf{y}'$. These newly generated knots are weighted equally, so an unweighted smoothing spline model is used. Following \cite{greenSilverman1994}, with

\begin{itemize}
  \item $\mathbf{E}' = \left(E'_{qr}\right)_{1 \leq q, r, \leq m'}$ defined by $E'_{qr} = \eta_{4-d}\left(\mathbf{y}'_{q} - \mathbf{y}'_{r}\right)$.
  \item $\mathbf{R}' = \left(R'_{qp}\right)_{1 \leq q \leq m', \ 1 \leq p \leq d + 1}$ is defined by $R'_{qp} = \phi_p(\mathbf{y}'_{q})$
\end{itemize}
the coefficients that minimize $\left(\mathbf{x}'_t - \mathbf{E}'\mathbf{s}_t - \mathbf{R}'^{T}\mathbf{\alpha}_t\right)^{T}\left(\mathbf{x}'_t - \mathbf{E}'\mathbf{s}_t - \mathbf{R}'^{T}\mathbf{\alpha}_t\right) + \gamma\mathbf{s}_t^{T}\mathbf{E}'\mathbf{s}_t$ can be found by solving 
\begin{equation}
  \left(
  \begin{array}{cc}
    \mathbf{E}' + \lambda_{t}\mathbf{I} & \mathbf{R}'^{T} \\
    \mathbf{R}' & \mathbf{0}
  \end{array}
  \right)\left(
  \begin{array}{c}
    \mathbf{s}'_t \\
    \mathbf{\alpha}'_t
  \end{array}
  \right) = \left(
  \begin{array}{c}
    \mathbf{x}'_t \\
    \mathbf{0}
  \end{array}
  \right). \label{eq:7}
\end{equation}

The coefficients found at each time point $t$, $\mathbf{s'}_{t}$ and $\mathbf{\alpha'}_{t}$ are concatenated in matrix $\mathbf{\omega}_{t} = \left(\mathbf{s'}_{t}^T, \mathbf{\alpha'}_{t}^T\right)$. This coefficient matrix is then used as the output value in a cubic smoothing spline function using the time points as input values. This function takes the form
\begin{equation}
  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i\|t - t_i\|^{3} + \sum_{p=1}^{2}\tau_p \phi(t)_p. \label{eq:7}
\end{equation}
We define matrix $\mathbf{A}$ as $\mathbf{A}_{qr} = \|t_q - t_r\|^{3}$, matrix $\mathbf{T}$ as $\mathbf{T}_{pr} = \phi_p(t_r)$, $\mathbf{\delta} = \left(\delta_1, \dots, \delta_T\right)^{T}$, and $\mathbf{\tau} = \left(\tau_1, \tau_2\right)^{T}$. Then, the coefficients of $g_{\gamma}(t)$, given a value of $\gamma$, can be found by solving
\begin{equation}
  \left(
  \begin{array}{cc}
    \mathbf{A} + \gamma\mathbf{I} & \mathbf{T}^{T} \\
    \mathbf{T} & \mathbf{0}
  \end{array}
  \right)\left(
  \begin{array}{c}
    \mathbf{\delta} \\
    \mathbf{\tau}
  \end{array}
  \right) = \left(
  \begin{array}{c}
    \mathbf{\omega} \\
    \mathbf{0}
  \end{array}
  \right). \label{eq:8}
\end{equation}

As with the tuning process for $\lambda$, the optimal value of $\gamma$ can be identified using cross-validation, with either leave-one-out or $k$-folds cross-validation being appropriate depending on the number of time points available. With a proper estimate of the optimum $\gamma$, we may denote the coefficients estimated by $g(t)$ as $g(t) = \Omega(t) = \left(\mathbf{s}_{\gamma}(t)^{T}, \mathbf{\alpha}_{\gamma}(t)^{T}\right)^{T}$. Therefore, the varying-coefficient models have the following outputs:
\begin{equation}
  f(\mathbf{y}, t) = \sum_{k=1}^{m'}\mathbf{s}_{\gamma}(t)_l \eta_{4-d}\left(\|\mathbf{y} - \mathbf{y'}_k\|\right) + \sum_{p=1}^{d + 1}\mathbf{\alpha}_{\gamma}(t)_p \phi_p(\mathbf{y}). \label{eq:9}
\end{equation}
The fitting process for the varying-coefficient smoothing spline model is given formally as Algorithm \ref{alg:varying_spline}.

\begin{algorithm}
\caption{Varying-Coefficient Smoothing Spline Model}\label{alg:varying_spline}
  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{ijk}\right\}_{i=1, j = 1, k = 1}^{N, n_i, m_{ij}}$, $d$-dimensional input values $\left\{\mathbf{y}_{ijk}\right\}_{i=1, j=1, k=1}^{N, n_i, m_{ij}}$, weights $\left\{w_{ijk}\right\}_{i=1, j=1, k=1}^{N, n_i, m_{ij}}$, observation times $\left\{t_{ij}\right\}_{i=1, j=1}^{N, n_i}$, candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$.}
  \KwResult{Analytic formula of $\hat{f}(\mathbf{y}, t)$ and optimal tuning parameter $\hat{\gamma}$.}

\For{$t = 1, 2, \dots, T$} {
  Set $U_t$ equal to the number of images included in the data at time $t$\;
  \For{$l = 1, 2, \dots, L$} {
    Solve equation \ref{eq:5} to estimate $f_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}_t$, $\mathbf{y}_t$, and $w_t$\;
    \For{$u = 1, 2, \dots, U_t$} {
      Solve equation \ref{eq:5} to estimate $f^{(-u)}_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}^{(-u)}_t$, $\mathbf{y}^{(-u)}_t$, $w^{(-u)}_t$\;
      Estimate $MSE(\lambda_{l})_u$
    }
  }
}
\end{algorithm}
